Contextuality and Nonlocality in Discrete Quantum Theory

### 2016-02-10 DQC3.tex
"6.1 Gleason's Theorem in DQT" Updates new results...

The last part of the theorem is to prove that for any $\ket{\Psi}\in\ffzd{p^{2}},\dpr_{\Psi}\ne\hat{\pi}$.
This part can be established by counting states. For any $\ket{\Psi}\in\ffzd{p^{2}}$,
\begin{eqnarray*}
 &  & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\dpr_{\Psi}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|\\
 & = & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\ip{\Psi}{i}=0\right\} \right|\\
 & < & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\hat{\pi}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|
\end{eqnarray*}
$\QED$
\end{proof}
There are reasons why there is no Gleason's theorem in DQT. First,
given ${\cal P}$ in CQT, we search for a mixed state $\rho$ such
that ${\cal P}={\cal P}_{\rho}$, but given $\hat{\pi}$ in DQT, we
only search for a pure state $\ket{\Psi}$ such that $\dpr_{\Psi}=\hat{\pi}$.
Notice that there are much more mixed states than pure states so that
it is likely we cannot find any pure state $\ket{\Psi}$ satisfying
$\dpr_{\Psi}=\hat{\pi}$ in DQT. Although this reason is important,
theorem \ref{thm:GleasonDQT} still hold even if we define a mixed
state. Assume $\rho$ is a mixed state between $\ket{\Psi_{1}}$ and
$\ket{\Psi_{2}}$, then we still have 
\begin{eqnarray*}
 &  & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\dpr_{\rho}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|\\
 & \le & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\dpr_{\Psi_{1}}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|\\
 & < & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\hat{\pi}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|
\end{eqnarray*}
Therefore, no matter how we define the mixed state, theorem \ref{thm:GleasonDQT}
still hold, and there are some other reasons why theorem \ref{thm:GleasonDQT}
hold.

Second, compare the following arithmetic equation and Boolean equation.
\begin{eqnarray}
{\cal P}\left(\ket{0}\right)+{\cal P}\left(\ket{1}\right)+{\cal P}\left(\ket{2}\right) & = & 1\label{eq:arithmetic}\\
\hat{\pi}\left(\ket{0}\right)\vee\hat{\pi}\left(\ket{1}\right)\vee\hat{\pi}\left(\ket{2}\right) & = & \text{possible}\label{eq:Boolean}
\end{eqnarray}
When ${\cal P}\left(\ket{0}\right)=\frac{1}{2}$, we still have an
equality ${\cal P}\left(\ket{1}\right)+{\cal P}\left(\ket{2}\right)=1$
for ${\cal P}\left(\ket{1}\right)$ and ${\cal P}\left(\ket{2}\right)$
because $\left(\mathbb{R},0,+\right)$ is a group. In contrast, when
$\hat{\pi}\left(\ket{0}\right)=\text{possible}$, we know nothing
about $\hat{\pi}\left(\ket{1}\right)$ and $\hat{\pi}\left(\ket{2}\right)$
because $\left(\mathscr{L}_{2},\text{impossible},\vee\right)$ is
a semigroup without cancellative, i.e., $a\vee c=b\vee c$ does not
imply $a=b$.

If the problem is that the range of $\hat{\pi}$ is only a semigroup,
we may want to consider the range of $\hat{\pi}$ to be a group.
\begin{problem}
\label{thm:GleasonDQT-group}Is there a group $\left(\mathscr{L},0,\oplus\right)$
with a special element $1\in\mathscr{L}$ and $1\ne0$ satisfying
the following statement? 
\begin{itemize}
\item There is a way to define a mixed state $\rho$ and a corresponding
Born rule $\tilde{\pi}_{\rho}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
in a sense of equation (\ref{prob.mixed}).
\item For $d\ge3$, for any $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
satisfying 
\begin{eqnarray}
 &  & \textrm{For any vector }\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{ , }\hat{\pi}(\ket{i})=\hat{\pi}(\gamma\ket{i})\textrm{ . }\label{eq:parallel-DQT-1}\\
 &  & \textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ , }{\displaystyle \bigoplus_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}1\textrm{ .}\label{eq:total-probability-DQT-1}
\end{eqnarray}
There is a mixed state $\rho$ such that $\hat{\pi}=\tilde{\pi}_{\rho}$.
\end{itemize}
\end{problem}
When $d=3$ and $p=7$, we have verified that we no longer have the
freedom to assign impossible more than any reasonable probability
assignment. Therefore, we cannot find a similar counterexample for
problem \ref{thm:GleasonDQT-group}. However, even if there is a group
$\left(\mathscr{L},0,\oplus\right)$ satisfying problem \ref{thm:GleasonDQT-group},
the corresponding Born rule $\tilde{\pi}_{\rho}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
is not basis independent, i.e., we will not have 
\begin{equation}
\tilde{\pi}_{U\rho}\left(U\ket{i}\right)=\tilde{\pi}_{\rho}\left(\ket{i}\right)\label{eq:rotation-1}
\end{equation}
for all unitary matrix $U$, mixed state $\rho$, and pure state $\ket{i}\in\ffzd{p^{2}}$,
and the proof is similar to the one of the no-go theorem in section
\ref{sub:A-No-Go-Theorem}.

>>>>

The last part of the theorem is to prove that for any $\ket{\Psi}\in\ffzd{p^{2}},\dpr_{\Psi}\ne\hat{\pi}$.
This part can be established by counting states. For any $\ket{\Psi}\in\ffzd{p^{2}}$,
\begin{eqnarray*}
 &  & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\dpr_{\Psi}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|\\
 & = & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\ip{\Psi}{i}=0\right\} \right|\\
 & < & \left|\left\{ \ket{i}\in\ffzd{p^{2}}\middle|\hat{\pi}\left(\ket{i}\right)=\textrm{impossible}\right\} \right|
\end{eqnarray*}
$\QED$
\end{proof}
It is easy to identify the reasons why theorem \ref{thm:GleasonDQT}
is different from Gleason's theorem in CQT, but it is hard to find
the correct conditions to get the corresponding Gleason's theorem
in DQT. Given ${\cal P}$ in CQT, we search for a mixed state $\rho$
such that ${\cal P}={\cal P}_{\rho}$, but given $\hat{\pi}$ in DQT,
we only search for a pure state $\ket{\Psi}$ such that $\dpr_{\Psi}=\hat{\pi}$.
Notice that there are much more mixed states than pure states so that
it is likely we cannot find any pure state $\ket{\Psi}$ satisfying
$\dpr_{\Psi}=\hat{\pi}$ in DQT.

Lacking the notion of mixed state is not the only reason why theorem
\ref{thm:GleasonDQT} is hold. Even if we find a way to define a mixed
state, if $\rho$ is a mixed state between $\ket{\Psi_{1}}$ and $\ket{\Psi_{2}}$,
then we still have 
\[
\left|\dpr_{\rho}^{-1}\left(\left\{ \textrm{impossible}\right\} \right)\right|\le\left|\dpr_{\Psi_{1}}^{-1}\left(\left\{ \textrm{impossible}\right\} \right)\right|<\left|\hat{\pi}^{-1}\left(\left\{ \textrm{impossible}\right\} \right)\right|\textrm{ .}
\]
We can always pick $\hat{\pi}$ such that $\left|\hat{\pi}^{-1}\left(\left\{ \textrm{impossible}\right\} \right)\right|$
is larger because we have $x\lor\textrm{possible}=\textrm{possible}$
no matter whether $x$ is possible or impossible. More precisely,
compare the following arithmetic equation and Boolean equation.
\begin{eqnarray}
{\cal P}\left(\ket{0}\right)+{\cal P}\left(\ket{1}\right)+{\cal P}\left(\ket{2}\right) & = & 1\label{eq:arithmetic}\\
\hat{\pi}\left(\ket{0}\right)\vee\hat{\pi}\left(\ket{1}\right)\vee\hat{\pi}\left(\ket{2}\right) & = & \text{possible}\label{eq:Boolean}
\end{eqnarray}
When ${\cal P}\left(\ket{0}\right)=\frac{1}{2}$, we still have an
equality ${\cal P}\left(\ket{1}\right)+{\cal P}\left(\ket{2}\right)=1$
for ${\cal P}\left(\ket{1}\right)$ and ${\cal P}\left(\ket{2}\right)$
because $\left(\mathbb{R},0,+\right)$ is a group. In contrast, when
$\hat{\pi}\left(\ket{0}\right)=\text{possible}$, we know nothing
about $\hat{\pi}\left(\ket{1}\right)$ and $\hat{\pi}\left(\ket{2}\right)$
because $\left(\mathscr{L}_{2},\text{impossible},\vee\right)$ is
a commutative semigroup without cancellative, i.e., $a\vee c=b\vee c$
does not imply $a=b$.

If the problem is that the range of $\hat{\pi}$ is only a commutative
semigroup, we may want to consider $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
, and $\left(\mathscr{L},0,\uplus\right)$ is an abelian group. By
the fundamental theorem of abelian group~\cite{DummitFoote2004},
$\mathscr{L}$ must be isomorphic to $\mathbb{Z}_{d_{1}}\oplus\ldots\oplus\mathbb{Z}_{d_{k}}\oplus\mathbb{Z}^{b}$,
where $\mathbb{Z}_{d_{1}},\ldots,\mathbb{Z}_{d_{k}}$ are finite cyclic
subgroups, $\mathbb{Z}^{b}$ is the free abelian group with rank $b$,
and $\oplus$ is the direct sum of the abelian groups. Moreover, we
want $\left(\mathscr{L},0,\uplus\right)$ to have a sense of probability. 
\begin{problem}
\label{thm:GleasonDQT-group}Is there a group $\left(\mathscr{L},0,\oplus\right)$
with a special element $1\in\mathscr{L}$ and $1\ne0$ satisfying
the following statement? 
\begin{itemize}
\item There is a way to define a mixed state $\rho$ and a corresponding
Born rule $\tilde{\pi}_{\rho}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
in a sense of equation (\ref{prob.mixed}).
\item For $d\ge3$, for any $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
satisfying 
\begin{eqnarray}
 &  & \textrm{For any vector }\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{ , }\hat{\pi}(\ket{i})=\hat{\pi}(\gamma\ket{i})\textrm{ . }\label{eq:parallel-DQT-1}\\
 &  & \textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ , }{\displaystyle \bigoplus_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}1\textrm{ .}\label{eq:total-probability-DQT-1}
\end{eqnarray}
There is a mixed state $\rho$ such that $\hat{\pi}=\tilde{\pi}_{\rho}$.
\end{itemize}
\end{problem}
When $d=3$ and $p=7$, we have verified that we no longer have the
freedom to assign impossible more than any reasonable probability
assignment. Therefore, we cannot find a similar counterexample for
problem \ref{thm:GleasonDQT-group}. However, even if there is a group
$\left(\mathscr{L},0,\oplus\right)$ satisfying problem \ref{thm:GleasonDQT-group},
the corresponding Born rule $\tilde{\pi}_{\rho}:\ffzd{p^{2}}\rightarrow\mathscr{L}$
is not basis independent, i.e., we will not have 
\begin{equation}
\tilde{\pi}_{U\rho}\left(U\ket{i}\right)=\tilde{\pi}_{\rho}\left(\ket{i}\right)\label{eq:rotation-1}
\end{equation}
for all unitary matrix $U$, mixed state $\rho$, and pure state $\ket{i}\in\ffzd{p^{2}}$,
and the proof is similar to the one of the no-go theorem in section
\ref{sub:A-No-Go-Theorem}.
